---
title: "MyCalc Documentation"
subtitle: "Developing Data Products Project"
author: "Russ Bjork"
date: "1/2/2021"
output: html_document
runtime: shiny
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
 
##    

  
This R Markdown document is made interactive using Shiny. Unlike the more traditional  
workflow of creating static reports, you can now create documents that allow your readers   
to change the assumptions underlying your analysis and see the results immediately.   

To learn more, see [Interactive Documents](http://rmarkdown.rstudio.com/authoring_shiny.html).  

## MyCalc Overview

The MyCalc application is a very simple calculator. The calculator provides the ability  
to perform calculations on selected numeric values.  The user selected values will be   
utilized in a simple calculation and the results will be returned to the user interface.

## MyCalc Usage

When the MyCalc application is rendered, a simple user interface will be presented.    
Start by first selecting the first numeric value from the 'select value 1' dropdown.    
The next step is to select the second numeric value from the 'select value 2' dropdown.    
After selecting values 1 and 2, Use the 'select computation method' dropdown to choose the   
preferred mathematical computation.  Finally, click on the 'Compute' button to run the computation.  
All values are reactively displayed in the main output panel as selections are made.    
Click the 'Clear' button to clear the computation output.  Enjoy!!!  

## MyCalc Embedded Application

The MyCalc application is embedded below.  This is a direct link to the published   
application on the RStudio Shiny Apps IO server:  

```{r, echo=FALSE}
knitr::include_app("https://russell-bjork.shinyapps.io/MyCalc/", height = "400px")
```





