---
title: "Statistical Inference - Project"
subtitle: "Part 2: Basic inferential data analysis"
author: "Russ Bjork"
date: "12/13/2020"
output: pdf_document
---

<body style="background-color:LightSkyBlue;">

# Overview:
Part 2 of the Statistical Inference project will focus on analyzing the ToothGrowth dataset.  This dataset is made of 60 observations of the tooth growth of Guinea pigs and the effect of a dose of vitamin C on the growth. The delivery method is a two-values factor variable : OJ (Orange Juice) and VC (Ascorbi Acid, as a form of Vitamin C). The dose given contains three different values : 0.5, 1, and 2 mg/day. The measures are done on 10 Guinea pigs.  

## Environment setup and configuration
```{r echo=TRUE}
## 
## Statistical Inference - Project
##
library(datasets)
library(dplyr)
library(ggplot2)
library(ggpubr)
library(knitr)
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_chunk$set(fig.width=12, fig.height=10)
```

# Load and Explore the Data:
Load the dataframe and review the data layout and date types  
```{r echo=TRUE}
str(ToothGrowth)
data(ToothGrowth)
head(ToothGrowth)
```
Check if there are NA values and display the summary information for this dataset  
```{r echo=TRUE}
sum(!complete.cases(ToothGrowth))
```

# Basic Summary of the Data and Plotting Key Variables:
Process a summary of the ToothGrowth data  
```{r echo=TRUE}
summary(ToothGrowth)
ToothGrowth$dose <- as.factor(ToothGrowth$dose)
ggplot(ToothGrowth, aes(x=dose, y=len,fill=dose)) +  
    geom_boxplot() +  
    ggtitle("Plot 1: Dispersion of tooth growth by dose") + 
    xlab("dose in mg") + ylab("tooth length")
```
The summary data shows that the higher the dose is, the longer the teeth are. It is also observed that for a dose of 1 mg, the mean is nearly twice that of a dose of 0.5 mg. The progression is observed to be lower when the dose is 2 mg. The position of the boxes are really different; this is a first clue for suggesting that the tooth length depends a lot of the dose.  

The following plot shows the influence of the delivery method.  
```{r echo=TRUE}
ggplot(ToothGrowth, aes(x=supp, y=len,fill=supp)) +  
    geom_boxplot() +  
    ggtitle("Plot 2: Dispersion of tooth growth by delivery method") + 
    xlab("delivery method") + ylab("tooth length")
```
The boxes are quite similar. However, the median is much higher for Orange Juice than for Vitamin C; that means that for Orange Juice, the high values are observed more frequently than the low values, in comparison with Vitamin C.  

# Confidence Intervals/Hypothesis Tests Comparing Tooth Growth by Supp and Dose:
A T test is used to determine if the delivery mode has an influence on the tooth growth. The nul hypothesis H0 could be formulated as follows:  
**H0 : The delivery mode of Vitamin C does not have any influence on the tooth growth**
```{r echo=TRUE}
dose <- ToothGrowth$dose
supp <- ToothGrowth$supp
len <-  ToothGrowth$len

t.test(len[supp == "VC"],len[supp == "OJ"], paired=FALSE)
```
The T test shows the following:    

- The p-value is 0.06, i.e. nearly the significance level lambda = 0.05. So we do not reject the null hypothesis,  
but as 0.05 <= 0.05 <= p-value <= 0.1 <= 0.1, it is not clearly obvious that we can reject the null hypothesis.  
- The confidence interval contains 0, so the test is not really significant.

The next T test is generated to test the influence of the dose on the tooth growth  
```{r echo=TRUE}
t.test(len[dose == 0.5],len[dose == 1], paired=FALSE)
```
This test shows that the quantity of Vitamin C has an impact on the tooth growth:    

- The p-value is nearly 0, so we can obviously reject H0  
- The confidence interval does not contain 0.  

An identical conclusion can be taken comparing the length of dose = 1 and dose = 2.    
This could already be detected from the boxplot above.    

# Conclusions and Assumptions:
- The dose of Vitamin C is clearly a factor of growth of the teeth for Guinea pigs.  
- The delivery mode (Ascorbic Acid or Orange Juice) does not have any obvious impact on the teeth growth for Guinea pigs.  
